---
title: "DSC532-Regression-EDA"
author: "Leonidas Ioannou, Ioustina Harasim, Fedor Turchenko"
date: "2023-04-23"
output: html_document
---

```{r setup, message=FALSE}
library(ggplot2)
library(tidyverse)
library(gridExtra)
library(corrplot)
library(pivottabler)
options(repr.plot.width = 8, repr.plot.height = 6, repr.plot.res = 200)
```

# Preprocessing
```{r}
# Import the data
d1=read.csv("student-mat.csv",sep=";",header=TRUE)
d2=read.csv("student-por.csv",sep=";",header=TRUE)

# Merge
df=merge(
  d1,
  d2[c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet")],
  by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet")
)

print(nrow(df)) # 382 students
```


```{r}
# Column names
colnames(df)
```

```{r}
# Dimension of the dataframe
dim(df)
```

```{r}
# Find columns with missing values
colSums(is.na(df))
```

```{r}
# Find duplicates
cat(round((sum(duplicated(df))/nrow(df))*100,3),'%') #percentage of duplicates
```

```{r}
sum(df$G3 == 0)
```

```{r}
df_num <- data.frame(df$age, df$failures, df$absences, df$G1, df$G2, df$G3)

# Finding outliers using IQR
# If a value is less than Q1-1.5IQR or greater than Q3+1.5IQR, it's considered as an outlier
find_outliers <- function(x){
  Q1 <- quantile(x, .25)
  Q3 <- quantile(x, .75)
  IQR <- IQR(x)
  Tmin = Q1 - 1.5*IQR
  Tmax = Q3 + 1.5*IQR
  num <- length(x[which(x < Tmin | x > Tmax)])
  num
}

outliers=c()
for (i in 1:ncol(df_num)){
  outliers[i]=find_outliers(df_num[1:nrow(df_num),i])
}
outliers = as.data.frame(round(((outliers/nrow(df_num))*100),2)) # Percentage of outliers for each feature
outliers$col = colnames(df_num) # Add the name of each column
outliers
```

# Data Exploration and visualization
# General Plots
```{r}
# Plot of the student's school
schoolplot <- df %>%
  count(school) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    school = as.factor(school)
  )

ggplot(schoolplot, aes(x = school , y = n, fill = school))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="School", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of the student's gender
genderplot <- df %>%
  count(sex) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    sex = as.factor(sex)
  )

ggplot(genderplot, aes(x = sex , y = n, fill = sex))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Gender", y="Count")  # Labels for x,y axes
```

```{r}
# Summary statistics of Age
summary(df$age)

p1<-ggplot(df,aes(age))+geom_histogram(fill="lightblue", color='black', bins=30) 
p2<-ggplot(df,aes(age))+geom_boxplot(fill="lightblue")
grid.arrange(p1,p2)
```

```{r}
# Plot of the student's address
# U - urban, R - rural
addressplot <- df %>%
  count(address) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    address = as.factor(address)
  )

ggplot(addressplot, aes(x = address , y = n, fill = address))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Address", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of the student's family size
# LE3 - less or equal to 3 or GT3 - greater than 3
famsizeplot <- df %>%
  count(famsize) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    famsize = as.factor(famsize)
  )

ggplot(famsizeplot, aes(x = famsize , y = n, fill = famsize))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Family Size", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of the student's parent's status
#  T - living together or A - apart
pstatusplot <- df %>%
  count(Pstatus) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    Pstatus = as.factor(Pstatus)
  )

ggplot(pstatusplot, aes(x = Pstatus , y = n, fill = Pstatus))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Parent's Cohabitation status", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of the student's mother's education
meduplot <- df %>%
  count(Medu) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    Medu = as.factor(Medu)
  )

ggplot(meduplot, aes(x = Medu , y = n, fill = Medu))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Mother's Education", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of the student's father's education
feduplot <- df %>%
  count(Fedu) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    Fedu = as.factor(Fedu)
  )

ggplot(feduplot, aes(x = Fedu , y = n, fill = Fedu))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
   theme(legend.position="none") +
  labs(x="Father's Education", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of the student's mother's job
mjobplot <- df %>%
  count(Mjob) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    Mjob = as.factor(Mjob)
  )

ggplot(mjobplot, aes(x = reorder(Mjob, n), y = n, fill = Mjob, n))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Mother's Job", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of the student's father's job
fjobplot <- df %>%
  count(Fjob) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    Fjob = as.factor(Fjob)
  )

ggplot(fjobplot, aes(x = reorder(Fjob, n), y = n, fill = Fjob, n))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Father's Job", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of reason to choose this school
reasonplot <- df %>%
  count(reason) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    reason = as.factor(reason)
  )

ggplot(reasonplot, aes(x = reorder(reason, n), y = n, fill = reason, n))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Reason", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's guardian
guardianplot <- df %>%
  count(guardian) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    guardian = as.factor(guardian)
  )

ggplot(guardianplot, aes(x = reorder(guardian, n), y = n, fill = guardian, n))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Guardian", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's traveltime
# 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
traveltimeplot <- df %>%
  count(traveltime) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    traveltime = as.factor(traveltime)
  )

  ggplot(traveltimeplot, aes(x = traveltime , y = n, fill = traveltime))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Travel Time", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's study time
# 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours
studytimeplot <- df %>%
  count(studytime) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    studytime = as.factor(studytime)
  )

ggplot(studytimeplot, aes(x = studytime , y = n, fill = studytime))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Study Time", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's failures
#  if 1<=n<3, else 4
failuresplot <- df %>%
  count(failures) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    failures = as.factor(failures)
  )

ggplot(failuresplot, aes(x = failures , y = n, fill = failures))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Failures", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's extra educational support
schoolsupplot <- df %>%
  count(schoolsup) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    schoolsup = as.factor(schoolsup)
  )

ggplot(schoolsupplot, aes(x = schoolsup , y = n, fill = schoolsup))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Educational Support", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's family educational support
familysupplot <- df %>%
  count(famsup) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    famsup = as.factor(famsup)
  )

ggplot(familysupplot, aes(x = famsup , y = n, fill = famsup))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Family Educational Support", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's extra paid classes 
paidplot <- df %>%
  count(paid) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    paid = as.factor(paid)
  )

ggplot(paidplot, aes(x = paid , y = n, fill = paid))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Paid Classes", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's activities 
activitiesplot <- df %>%
  count(activities) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    activities = as.factor(activities)
  )

ggplot(activitiesplot, aes(x = activities , y = n, fill = activities))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Activities", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's attendance in nursery school 
nurseryplot <- df %>%
  count(nursery) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    nursery = as.factor(nursery)
  )

ggplot(nurseryplot, aes(x = nursery , y = n, fill = nursery))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Attend Nursery School", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's - higher education
higherplot <- df %>%
  count(higher) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    higher = as.factor(higher)
  )

ggplot(higherplot, aes(x = higher , y = n, fill = higher))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Higher Education", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's internet access at home 
internetplot <- df %>%
  count(internet) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    internet = as.factor(internet)
  )

ggplot(internetplot, aes(x = internet , y = n, fill = internet))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Internet Access", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's romantic relationship
relantionshipplot <- df %>%
  count(romantic) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    romantic = as.factor(romantic)
  )

ggplot(relantionshipplot, aes(x = romantic , y = n, fill = romantic))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Romantic Relantionship", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's quality of family relationship
# 1 - very bad to 5 - excellent
famrelplot <- df %>%
  count(famrel) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    famrel = as.factor(famrel)
  )

ggplot(famrelplot, aes(x = famrel , y = n, fill = famrel))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Quality of Family Relantionship", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's free time
# 1 - very low to 5 - very high
freetimeplot <- df %>%
  count(freetime) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    freetime = as.factor(freetime)
  )

ggplot(freetimeplot, aes(x = freetime , y = n, fill = freetime))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Free Time", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's going out with friends
# 1 - very low to 5 - very high
gooutplot <- df %>%
  count(goout) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    goout = as.factor(goout)
  )

ggplot(gooutplot, aes(x = goout , y = n, fill = goout))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Going Out", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's workday alcohol consumption 
# 1 - very low to 5 - very high
dalcplot <- df %>%
  count(Dalc) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    Dalc = as.factor(Dalc)
  )

ggplot(dalcplot, aes(x = Dalc , y = n, fill = Dalc))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Workday Alcohol Consumption", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's weekend alcohol consumption 
walcplot <- df %>%
  count(Walc) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    Walc = as.factor(Walc)
  )

ggplot(walcplot, aes(x = Walc , y = n, fill = Walc))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Weekend Alcohol Consumption", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's current health
# 1 - very bad to 5 - very good
healthplot <- df %>%
  count(health) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    health = as.factor(health)
  )

ggplot(healthplot, aes(x = health , y = n, fill = health))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="Health Status", y="Count")  # Labels for x,y axes
```

```{r}
# Plot of student's number of school absences
summary(df$absences)

p1<-ggplot(df,aes(absences))+geom_histogram(fill="#c19ce3", color='black', bins=30) 
p2<-ggplot(df,aes(absences))+geom_boxplot(fill="#c19ce3")
grid.arrange(p1,p2)
```

```{r}
# Plot of student's first period grade
summary(df$G1)

p3<-ggplot(df,aes(G1))+geom_histogram(fill="#e2ec39", color='black', bins=30) 
p4<-ggplot(df,aes(G1))+geom_boxplot(fill="#e2ec39")
grid.arrange(p3,p4)
```

```{r}
# Plot of student's second period grade
summary(df$G2)

p5<-ggplot(df,aes(G2))+geom_histogram(fill="#81a2f9", color='black', bins=30) 
p6<-ggplot(df,aes(G2))+geom_boxplot(fill="#81a2f9")
grid.arrange(p5,p6)
```

```{r}
# Plot of student's final grade
summary(df$G3)
var(df$G3)

p7<-ggplot(df,aes(G3))+geom_histogram(fill="#92ebd6", color='black', bins=30) 
p8<-ggplot(df,aes(G3))+geom_boxplot(fill="#92ebd6")
grid.arrange(p7,p8)
```

# Check how other variables affects the target variable G3
```{r}
# Plot of the final grade by school
ggplot(df, mapping = aes(x = school, y = G3, fill = school)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by School")
```

```{r}
# Plot of the final grade by gender
ggplot(df, mapping = aes(x = sex, y = G3, fill = sex)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Gender")
```

```{r}
# Plot of the final grade by address
ggplot(df, mapping = aes(x = address, y = G3, fill = address)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Address")
```

```{r}
# Plot of the final grade by family size
ggplot(df, mapping = aes(x = famsize, y = G3, fill = famsize)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Family Size")
```

```{r}
# Plot of the final grade by parent's status
ggplot(df, mapping = aes(x = Pstatus, y = G3, fill = Pstatus)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Parent's Status")
```

```{r}
# Plot of the final grade by mother's education
ggplot(df, mapping = aes(x = Medu, y = G3, fill = as.character(Medu))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Mother's Education", fill='Medu')
```


```{r}
# Plot of the final grade by father's education
ggplot(df, mapping = aes(x = Fedu, y = G3, fill = as.character(Fedu))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Father's Education", fill='Fedu')
```

```{r}
# Plot of the final grade by mother's job
ggplot(df, mapping = aes(x = Mjob, y = G3, fill = as.character(Mjob))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Mother's Job", fill='Mjob')
```

```{r}
# Plot of the final grade by father's job
ggplot(df, mapping = aes(x = Fjob, y = G3, fill = as.character(Fjob))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Father's Job", fill='Fjob')
```

```{r}
# Plot of the final grade by reason to chose this school
ggplot(df, mapping = aes(x = reason, y = G3, fill = reason)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Reason")
```

```{r}
# Plot of the final grade by guardian
ggplot(df, mapping = aes(x = guardian, y = G3, fill = guardian)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Guardian")
```

```{r}
# Plot of the final grade by travel time
ggplot(df, mapping = aes(x = traveltime, y = G3, fill = as.character(traveltime))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Travel Time", fill='Travel Time')
```

```{r}
# Plot of the final grade by study time
ggplot(df, mapping = aes(x = studytime, y = G3, fill = as.character(studytime))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Study Time", fill='Study Time')
```

```{r}
# Plot of the final grade by failures
ggplot(df, mapping = aes(x = failures, y = G3, fill = as.character(failures))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Failures", fill='Failures')
```

```{r}
# Plot of the final grade by school support
ggplot(df, mapping = aes(x = schoolsup, y = G3, fill = schoolsup)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by School Support")
```

```{r}
# Plot of the final grade by family educational support
ggplot(df, mapping = aes(x = famsup, y = G3, fill = famsup)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Family Educational Support")
```

```{r}
# Plot of the final grade by paid classes
ggplot(df, mapping = aes(x = paid, y = G3, fill = paid)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Paid Classes")
```

```{r}
# Plot of the final grade by activities
ggplot(df, mapping = aes(x = activities, y = G3, fill = activities)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Activities")
```

```{r}
# Plot of the final grade by attendance in nursery school
ggplot(df, mapping = aes(x = nursery, y = G3, fill = nursery)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Nursery School")
```

```{r}
# Plot of the final grade by higher education
ggplot(df, mapping = aes(x = higher, y = G3, fill = higher)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Higher Education")

# students who are planning on a higher education perform better than those who do not
```

```{r}
# Plot of the final grade by internet access
ggplot(df, mapping = aes(x = internet, y = G3, fill = internet)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Internet Access")
```

```{r}
# Plot of the final grade by romantic relationship
ggplot(df, mapping = aes(x = romantic, y = G3, fill = romantic)) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Romantic Relantionship")
```

```{r}
# Plot of the final grade by quality of family relantionship
ggplot(df, mapping = aes(x = famrel, y = G3, fill = as.character(famrel))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Quality of Family Relationship", fill='FamRel')
```

```{r}
# Plot of the final grade by free time
ggplot(df, mapping = aes(x = freetime, y = G3, fill = as.character(freetime))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Free Time", fill='Free Time')
```

```{r}
# Plot of the final grade by going out with friends
ggplot(df, mapping = aes(x = goout, y = G3, fill = as.character(goout))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Going out", fill='Goout')
```

```{r}
# Plot of the final grade by workday alcohol consumption 
ggplot(df, mapping = aes(x = Dalc, y = G3, fill = as.character(Dalc))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by workday alcohol consumption ", fill='Dalc')
```

```{r}
# Plot of the final grade by weekend alcohol consumption
ggplot(df, mapping = aes(x = Walc, y = G3, fill = as.character(Walc))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by weekend alcohol consumption ", fill='Walc')
```

```{r}
# Plot of the final grade by health status
ggplot(df, mapping = aes(x = health, y = G3, fill = as.character(health))) + 
  geom_boxplot(outlier.colour="black") +
  labs(title="Final Grade by Health Status", fill='health')
```

```{r}
# Plot of the final grade by absences
ggplot(df,aes(x=absences,y=G3)) + geom_point()
```

```{r}
# Plot of G3 vs G1,G2
p9<-ggplot(df,aes(x=G1,y=G3)) + geom_point() + geom_smooth(method = 'lm')
p10<-ggplot(df,aes(x=G2,y=G3)) + geom_point() + geom_smooth(method = 'lm')
grid.arrange(p9,p10)
```

# Ordinal Encoding
```{r}
# Encoding of the variables
encode_ordinal <- function(x, order = unique(x)) {
  x <- as.numeric(factor(x, levels = order, exclude = NULL))
  x
}

df_encoded <- df
df_encoded$school <- encode_ordinal(df$school)
df_encoded$sex <- encode_ordinal(df$sex)
df_encoded$address <- encode_ordinal(df$address)
df_encoded$famsize <- encode_ordinal(df$famsize)
df_encoded$Pstatus <- encode_ordinal(df$Pstatus)
df_encoded$Mjob <- encode_ordinal(df$Mjob)
df_encoded$Fjob <- encode_ordinal(df$Fjob)
df_encoded$reason <- encode_ordinal(df$reason)
df_encoded$guardian <- encode_ordinal(df$guardian)
df_encoded$schoolsup <- encode_ordinal(df$schoolsup)
df_encoded$famsup <- encode_ordinal(df$famsup)
df_encoded$paid <- encode_ordinal(df$paid)
df_encoded$activities <- encode_ordinal(df$activities)
df_encoded$nursery <- encode_ordinal(df$nursery)
df_encoded$higher <- encode_ordinal(df$higher)
df_encoded$internet <- encode_ordinal(df$internet)
df_encoded$romantic <- encode_ordinal(df$romantic)
```

# Correlation Matrix
```{r}
# Correlation among variables
corrplot(cor(df_encoded), tl.cex=0.5)
```

# Impact of parents on the performance of students
```{r}
df %>% group_by(Medu) %>% summarize('g1'=mean(G1),'g2'=mean(G2),'g3'=mean(G3),n=n())
df %>% group_by(Fedu) %>% summarize('g1'=mean(G1),'g2'=mean(G2),'g3'=mean(G3),n=n()) 
```

```{r}
df %>% group_by(Mjob) %>% summarize('g1'=mean(G1),'g2'=mean(G2),'g3'=mean(G3),n=n()) %>% 
  ggplot(aes(x=Mjob))+geom_line(aes(y=g1),color='blue',group=1,lwd=2)+
  geom_line(aes(y=g2),color='green',group=1,lwd=2)+geom_line(aes(y=g3),color='red',group=1,lwd=2)+
  labs(title="Was mother's job helpful?",y='Performance',subtitle='G1:blue, G2:green, G3:red')+
  theme(text=element_text(size=20)) 
```

```{r}
df %>% group_by(Fjob) %>% summarize('g1'=mean(G1),'g2'=mean(G2),'g3'=mean(G3),n=n()) %>% 
  ggplot(aes(x=Fjob))+geom_line(aes(y=g1),color='blue',group=1,lwd=2)+
  geom_line(aes(y=g2),color='green',group=1,lwd=2)+geom_line(aes(y=g3),color='red',group=1,lwd=2)+
  labs(title="Was father's job helpful?",y='Performance',subtitle='G1:blue, G2:green, G3:red')+
  theme(text=element_text(size=20))
```

```{r}
df %>% group_by(Medu) %>% summarize('g1'=mean(G1),'g2'=mean(G2),'g3'=mean(G3),n=n()) %>% 
  ggplot(aes(x=Medu))+geom_line(aes(y=g1),color='blue',group=1,lwd=2)+
  geom_line(aes(y=g2),color='green',group=1,lwd=2)+geom_line(aes(y=g3),color='red',group=1,lwd=2)+
  labs(title="Was mother's education helpful?",y='Performance',subtitle='G1:blue, G2:green, G3:red')+
  theme(text=element_text(size=20)) 

```


```{r}
df %>% group_by(Fedu) %>% summarize('g1'=mean(G1),'g2'=mean(G2),'g3'=mean(G3),n=n()) %>% 
  ggplot(aes(x=Fedu))+geom_line(aes(y=g1),color='blue',group=1,lwd=2)+
  geom_line(aes(y=g2),color='green',group=1,lwd=2)+geom_line(aes(y=g3),color='red',group=1,lwd=2)+
  labs(title="Was father's education helpful?",y='Performance',subtitle='G1:blue, G2:green, G3:red')+
  theme(text=element_text(size=20))

```
# Combine Female and Male columns
```{r}
df$MplusFJob <- paste(df$Mjob, "+", df$Fjob)
df$MplusFedu <- paste(df$Medu + df$Fedu)
```

```{r}
df %>% group_by(MplusFedu) %>% summarize('g1'=mean(G1),'g2'=mean(G2),'g3'=mean(G3),n=n())
```

```{r}
df %>% group_by(MplusFJob) %>% summarize('g1'=mean(G1),'g2'=mean(G2),'g3'=mean(G3),n=n()) %>% 
  ggplot(aes(x=MplusFJob))+geom_line(aes(y=g1),color='blue',group=1,lwd=2)+
  geom_line(aes(y=g2),color='green',group=1,lwd=2)+geom_line(aes(y=g3),color='red',group=1,lwd=2)+
  labs(title="Was mother's education helpful?",y='Performance',subtitle='G1:blue, G2:green, G3:red')+
  theme(text=element_text(size=5), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

```{r}
df %>% group_by(MplusFedu) %>% summarize('g1'=mean(G1),'g2'=mean(G2),'g3'=mean(G3),n=n()) %>% 
  ggplot(aes(x=MplusFedu))+geom_line(aes(y=g1),color='blue',group=1,lwd=2)+
  geom_line(aes(y=g2),color='green',group=1,lwd=2)+geom_line(aes(y=g3),color='red',group=1,lwd=2)+
  labs(title="Was mother's education helpful?",y='Performance',subtitle='G1:blue, G2:green, G3:red')+
  theme(text=element_text(size=20)) 
```

# Evidence from correlation matrix
```{r}
# Address and travel time seem to be moderately correlated
# However, these variables are categorical, so it is better to check the relationship between them through Chi-squared test
df_chisq <- PivotTable$new()
df_chisq$addData(df)
df_chisq$addColumnDataGroups("address")
df_chisq$addRowDataGroups("traveltime")
df_chisq$defineCalculation(calculationName="count", summariseExpression="n()")
df_chisq$evaluatePivot()
df_chisq <- df_chisq$asDataFrame()
df_chisq <- df_chisq[, !names(df_chisq) %in% c("Total")]
df_chisq <- head(df_chisq, -1)
df_chisq
```

```{r}
chisq <- chisq.test(df_chisq)
chisq
```

Since there was a warning of Chi-squared approximation to potentially be incorrect, some sizes of counts are too small, and so the Fisher's exact test will be used for more accurate results

```{r}
fisher <- fisher.test(df_chisq)
fisher
```

H0: There is no dependency between type of area (rural/urban) and time it takes to go to school. Since the p-value is lower than 0.05, there is enough evidence to reject the null hypothesis at 95% confidence level. Therefore, there is a relationship between address and travel time, and thus these variables might be collinear.

Both variables represent quite similar concept of time that it takes to go to school. Seems logical to leave only one of these variables in the model

```{r}
# `Failures` and `higher` variables time seem to be moderately correlated as well
# However, these variables are also categorical, so it is a more accurate approach to check the relationship between them with Chi-squared test
df_chisq <- PivotTable$new()
df_chisq$addData(df)
df_chisq$addColumnDataGroups("higher")
df_chisq$addRowDataGroups("failures")
df_chisq$defineCalculation(calculationName="count", summariseExpression="n()")
df_chisq$evaluatePivot()
df_chisq <- df_chisq$asDataFrame()
df_chisq <- df_chisq[, !names(df_chisq) %in% c("Total")]
df_chisq <- head(df_chisq, -1)
df_chisq
```

```{r}
chisq <- chisq.test(df_chisq)
chisq
```

Since there was a warning of Chi-squared approximation to potentially be incorrect, some sizes of counts are too small, and so the Fisher's exact test will be used for more accurate results

```{r}
fisher <- fisher.test(df_chisq)
fisher
```

H0: There is no dependency between willingness to take higher education and failures of classes. Since the p-value is lower than 0.05, there is enough evidence to reject the null hypothesis at 95% confidence level. Therefore, there is a relationship between higher and failures variables, and thus these variables might be collinear.

The majority of students want to take higher education and never failed a class, so, in fact, running a test was not a necessity here. Most probably, both of these variables will not provide much information, and so both could be dropped

```{r}
# Workday and weekend alcohol consumption could be combined into a single predictor
# It will still represent how much alcohol students consume but will remove unnecessary collinearity
df$alc <- df$Walc+df$Dalc
```

```{r}
# Alcohol consumption level variables were correlated with going out – let's check this relationship with chi-squared test
# However, these variables are also categorical, so it is a more accurate approach to check the relationship between them with Chi-squared test
df_chisq <- PivotTable$new()
df_chisq$addData(df)
df_chisq$addColumnDataGroups("goout")
df_chisq$addRowDataGroups("alc")
df_chisq$defineCalculation(calculationName="count", summariseExpression="n()")
df_chisq$evaluatePivot()
df_chisq <- df_chisq$asDataFrame()
df_chisq <- df_chisq[, !names(df_chisq) %in% c("Total")]
df_chisq <- head(df_chisq, -1)
df_chisq[is.na(df_chisq)] <- 0
df_chisq
```

```{r}
chisq <- chisq.test(df_chisq)
chisq
```
Since there was a warning of Chi-squared approximation to potentially be incorrect, some sizes of counts are too small, and so the Fisher's exact test will be used for more accurate results

```{r}
fisher <- fisher.test(df_chisq, simulate.p.value = TRUE) # simulate p-value due to very low frequencies in certain classes
fisher
```

H0: There is no dependency between level of alcohol consumption and frequency of going out with friends. Since the p-value is lower than 0.05, there is enough evidence to reject the null hypothesis at 95% confidence level. Therefore, there is a relationship between goout and alc variables, and thus these variables might be collinear.

It is quite clear that the more frequently students hang out with friends, the more alcohol they consume. Since going out is a bit broader concept, which also involves other activities than drinking alcohol, we can drop alcohol consumption variable to reduce potential collinearity

# Evidence from plots
Mother's and father's jobs are in majority represented by 'other' class
It does not provide much evidence in terms of explaining which exact parents' job could force students to study harder
However, one can assume that:

If one of the parents works in a healthcare, then they spent more years when getting their education, so they can positively influence child's attitude to studying
If one of the parents works as a teacher, then they better know how to motivate their child to study harder, explain some material in a more detail, and so the grades will be better

```{r}
df$parent_health_teacher = ifelse((df$Fjob=="teacher")|(df$Mjob=="teacher")|(df$Mjob=="health")|(df$Fjob=="health"), 1, 0)
```

```{r}
# Plot of the new variable
p_hplot <- df %>%
  count(parent_health_teacher) %>% # Count the number of occurrences
  mutate(
    perc = round(proportions(n) * 100, 1), # Percentage
    res = str_c(n, "(", perc, ")%"),
    parent_health_teacher = as.factor(parent_health_teacher)
  )

ggplot(p_hplot, aes(x = reorder(parent_health_teacher, n), y = n, fill = parent_health_teacher, n))  +
  geom_col() +
  geom_text(aes(label = res), colour = "black", vjust = -0.5) + 
  theme(legend.position="none") +
  labs(x="One of parent's job is healthcare or teaching", y="Count")  # Labels for x,y axes
```


# Possible Interactions
Interactions with sex variable could be one of the best in terms of explanation of why this interaction may occur naturally. Therefore, in this section we will check if some gender-based interaction terms could be used for our model

Sex and study time – to determine if the relationship between study time and grades differ among genders. We are interested in checking if at certain level of spent time on studying one of genders could outperform the other in terms of final score

```{r}
ggplot(df, aes(G3)) +
  geom_histogram(binwidth = 1.5) +
  facet_wrap(
    ~ df$studytime + df$sex,
    nrow=2,
    ncol=4,
    scale="free"
  ) +
  scale_x_continuous(
    limits=c(1, 20)
  ) +
  theme_light() +
  theme(plot.title=element_text(hjust = 0.5)) +
  xlab("Final score (0-20)") +
  ylab("Frequency") +
  ggtitle("Final Score Grouped by Weekly Study Time and Genders")
```
Distributions do not differ much for the classes 1-2 of study time. As for study time 3-4, there are too few male students within these groups, and the difference is still not that considerable

Next interaction to test is the sex and romantic one. While being in relationship, boys and girls may show varying attitude towards studying, so it may be possible that there could be a difference in finals score

```{r}
ggplot(df, aes(G3)) +
  geom_bar() +
  facet_wrap(
    ~ df$romantic + df$sex,
    nrow=2,
    ncol=2,
    scale="free"
  ) +
  scale_x_continuous(
    limits=c(1, 20)
  ) +
  theme_light() +
  theme(plot.title=element_text(hjust = 0.5)) +
  xlab("In a romantic relationship: yes/no") +
  ylab("Frequency") +
  ggtitle("Final Score Grouped by Being in Relationship and Genders")
```
Further test of possible interaction is devoted to sex and absences variables. We would like to investigate if there is a case when for representatives of one gender the number of absences has different relationship with the final grade than for another

In order to test this assumption, we will make a scatter plot of absences and sex, fit the linear model to visualize relationship. For a more accurate and representative plotting, there was used an approach of Poisson regression. Therefore, we would like to plot  𝑙𝑜𝑔(𝜆𝑖)  of G3 by number of absences.  𝜆𝑖  is unknown, so we approximate it by taking average value of final score at every value of number of absences and respective gender

```{r}
df_plot = df %>% group_by(absences, sex) %>% summarize("G3"=log(mean(G3)))
ggplot(
  data=df_plot,
  aes(
    x=absences,
    y=G3,
    color=sex
  )
) +
  geom_point() +
  geom_smooth(
    method="lm",
    formula=y ~ x,
    se=FALSE
  ) +
  xlab("Number of school absences") +
  ylab("Log empirical mean of final grade at number of absences")
```
For male students there is quite considerable negative linear relationship between number of school absences and final score. However, it is not that obvious for girls, which means that girls can actually achieve quite high score even at higher number of absences. This might be caused by the fact that girls may skip classes for a good reason, while boys tend to intentionally do it. Therefore, it could be useful to add an interaction  𝑠𝑒𝑥×𝑎𝑏𝑠𝑒𝑛𝑐𝑒𝑠  to the model

```{r}
colnames(df)
```

```{r}
encode_ordinal <- function(x, order = unique(x)) {
  x <- as.numeric(factor(x, levels = order, exclude = NULL))
  x
}

df$school <- encode_ordinal(df$school)
df$sex <- encode_ordinal(df$sex)
df$famsize <- encode_ordinal(df$famsize)
df$Pstatus <- encode_ordinal(df$Pstatus)
df$reason <- encode_ordinal(df$reason)
df$guardian <- encode_ordinal(df$guardian)
df$schoolsup <- encode_ordinal(df$schoolsup)
df$famsup <- encode_ordinal(df$famsup)
df$paid <- encode_ordinal(df$paid)
df$activities <- encode_ordinal(df$activities)
df$nursery <- encode_ordinal(df$nursery)
df$internet <- encode_ordinal(df$internet)
df$romantic <- encode_ordinal(df$romantic)
df$goout <- encode_ordinal(df$goout)
df$MplusFedu <- encode_ordinal(df$MplusFedu)
df$famrel <- encode_ordinal(df$famrel)
df$health <- encode_ordinal(df$health)
```
